---
title: "A brief analysis of Storm events data with NOAA information"
author: "Franco Israel Corona Hernandez"
date: "7/11/2021"
output: 
  html_document:
    keep_md: true
---

## Previus configs
```{r setup II, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(eval = TRUE, echo = TRUE)
#install.packages("dplyr")
#install.packages("stringdist")
library("dplyr")            ## load
library("stringdist")
library("ggplot2")            ## load


```

  * Done

## Synopsis

Climate is an important part of our lives and world so we better try to understand the most of it to know how it affect us. 
Here I present an analysis of the principal events recorded from 1950 to 2011 and the question we want to address are:

1.	Across the United States, which types of events (as indicated in the EVTYPE are most harmful with respect to population health?
2.	Across the United States, which types of events have the greatest economic consequences?

To answer this, we are provided of a data set from the Storm data events recorded by NOAA. 
First, we adjust the data to fit our purpose, this includes data transformations and calculations based on multiple resources and investigation. 
As this a critical part of our analysis this report is covered mostly of the adjustments and a brief review of the data and how it was transformed.

Finally, my results are presented as a quick view of the top events for each question. 

For further information and doubts regarding this report please contact me.

 

## Getting data
```{r gettingInicialFile}
final_file = "data.csv.bz2"
url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if(!file.exists(final_file)){
  download.file(url,final_file,quiet = FALSE,method = "curl")
}else{ 
  print(paste("File ",final_file,"already exists. Skipping download.",collapse = ""))
}

```
## Understanding data in file
 - Showing 10 instead of 1000 original lines loaded to keep more clean the output / report

```{r briefRaw}
file_raw <- readLines(con<-file(final_file),n=1000)
close(con)
head(file_raw,10)
```

This brief of file shows us that there is a comma separated file with headers, double quote as columns strings enclosure. 
Most of the data is numeric, at first sight 5 colulmns are posible string factors , it seems that null values are empty string, dates have date format 4/18/1950 and have a part of time in 0, time is in a INT format in 24 hrs format (1600)

## Loading and Processing the data

```{r loadDataR, cache=TRUE}
data_pp_1 <- read.table(file = final_file,sep = ",",header = TRUE,quote = "\"",na.strings = c("","NA"))

```

Loading seems correct now, a litte metrics in general about the dataset

```{r summary1}
summary(data_pp_1)
```

 + First values
 
```{r headFirstRows}

head(data_pp_1,10)

```

  + Last Values

```{r tailLastRows}
tail(data_pp_1,10)
```



Observations:  
  - Total Number of rows is **`r nrow(data_pp_1) `**, **`r ncol(data_pp_1) `** variables  
  - *F* column is almost empty on all the data.set  
  - *COUNTYENDN* is always NA  
  - *COUNTY_END* is always 0  
  - *Length*, *Width* are 0 at least at the 3rd quantile  
  - *BGN_TIME* has mixed values, integers and time strings  

* Using a different data.frame to keep the original for further checks

```{r dfClone}
data_pp_2<-data.frame()
data_pp_2<-data_pp_1
```

* Changing data types of dates: BGN_DATE,END_DATE, this for further manipulation and plot of the health damage by year

```{r transforms}

data_pp_2$BGN_DATE <- as.Date(data_pp_2$BGN_DATE,format="%m/%d/%Y %H:%M:%S")
data_pp_2$END_DATE <- as.Date(data_pp_2$END_DATE,format="%m/%d/%Y %H:%M:%S")
summary(data_pp_2[,c("BGN_DATE","END_DATE")])

```

Further transformations are included in the Questions section

## Questions

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

```

1.1 - I'm going to analyse the EVTYPE column

```{r eval=TRUE, include=TRUE}
evt <-
  data_pp_2 %>%
  group_by(EVTYPE) %>%
  summarize(total = n())
#evt
head(evt[order(-evt$total),],10)

```

Because there are `r length(evt)` different events, I decided that we could use some help (1). Getting a cleaner catalog of events, therefore we are going to match this by string distance. 

First some cleaning and data wrangling

```{r}
event_type_catalog <- read.csv2(file = "EVTYPE.csv",header = TRUE)
evt$leandata <- evt$EVTYPE
evt$leandata <- as.character(evt$leandata)
evt$leandata <- trimws(evt$leandata)
evt$leandata <- tolower(evt$leandata)
event_type_catalog$EVTYPE <- as.character(event_type_catalog$EVTYPE)
event_type_catalog$EVTYPE <- trimws(event_type_catalog$EVTYPE)
event_type_catalog$EVTYPE <- tolower(event_type_catalog$EVTYPE)
```

Quick look of the data.frames

```{r}
head(evt,10)
head(event_type_catalog,10)
```
How many rows by event type the original data.frame has, quick view

```{r eval=TRUE, include=TRUE}
evt2 <- 
  evt %>%
  group_by(leandata) %>%
  summarize(total = sum(total))

head(evt2[order(-evt2$total),],10)
```

Starting the process of getting the distances by jaro-winker method (closer to 0 is better)

```{r}

# Reclassification of EVTYPE to get a more accurate global classification

# Actual eventtype (dirty data)
uniquemodels <- unique(as.character(evt2$leandata))
# Catalog of NOAA to match
uniquemodels2 <- unique(as.character(event_type_catalog$EVTYPE)) 

# Calculation of the string distances by Jaro–Winkler
distancemodels <- stringdistmatrix(uniquemodels2,uniquemodels,method = "jw")

# setting colnames and rownames to identify which row and column correspond to each one distance
rownames(distancemodels) <- uniquemodels2
colnames(distancemodels) <- uniquemodels
#hc <- hclust(as.dist(distancemodels,upper = TRUE))
#plot(hc)
#rect.hclust(hc,k=20)
#list of minimun values (minimum values per column)
mins<-apply(distancemodels, 2, min)
#return list of the row number of the minimum value (most similar row vs columns)
minlist<-apply(distancemodels, 2, which.min)
#return list of matching values, minlist are the values of the NOAA catalog with is more likely to match the dirty data
matchwith<-uniquemodels2[minlist]

answer<-data.frame(uniquemodels, matchwith, mins)
maxDist<-0.30
answer_filter <- answer[(answer$mins<=maxDist)&(order(-answer$mins)),]
head(answer_filter[order(-answer_filter$mins),],10)

```
Getting in the **evt2** data.frame the values adjusted to the master catalog of event_types
```{r}
evt2_match <- merge(evt2,answer_filter,by.x=c("leandata"),by.y=c("uniquemodels"),all.x = TRUE)
head(evt2_match[order(-evt2_match$total),],10)
```

Lets do a quick analysis of the percentiles in where the data is null, separate the data in null and not null

```{r}
evt2_match$factor_matchwith<-0
evt2_match[complete.cases(evt2_match),'factor_matchwith'] <- 1

```


```{r}
evt3 <- 
  evt2_match %>%
  group_by(factor_matchwith) %>%
  summarize(total = sum(total), total_cn = n())

evt3[order(-evt3$total),]
```
At a **maxDist** of .30 we found a good proportion of data events  re-classified, `r (evt3[evt3$factor_matchwith==0,'total']/evt3[evt3$factor_matchwith==1,'total'])*100`% are misspelled events, summary or something that occur one time and not relevant

A little review of the data.frames I have  
**data_pp_2** has the full data, whit a few transformations  
**evt** has the relation between original *EVTYPE* column and *leandata* (the one used to measure distances)  
**evt2_match** has the final data.frame with a more clear classification (matchwith column)

```{r}
x0<-merge(evt,evt2_match,by.x=c("leandata"),by.y=c("leandata"),all.x = TRUE)
x<-merge(data_pp_2,x0,by.x=c("EVTYPE"),by.y=c("EVTYPE"),all.x = TRUE) 
#matchwith is the final event_type_colum
```

1.2 Checking the rest of the columns I discovered two columns, one of fatalities and other for injuries, for practical purposes we're going to break that down by year and if it was fatal o not, a plot would by helpful here.

```{r}
x_p1 <-
x %>%
  group_by(matchwith) %>%
  summarize(fatal = sum(FATALITIES), non_fatal = sum(INJURIES))
x_p1$total <- x_p1$fatal + x_p1$non_fatal
x_p1 <- x_p1[order(-x_p1$total),]
```

This **plot** to see over the years how injuries evolve

```{r}
x_p1_1 <-
x %>%
  group_by(year=format(BGN_DATE,'%Y')) %>%
  summarize(fatal = sum(FATALITIES), non_fatal = sum(INJURIES))
x_p1_1$total <- x_p1_1$fatal + x_p1_1$non_fatal
x_p1_1 <- x_p1_1[order(x_p1_1$year),]

# using a factor for fatalities would be more practical
g<-ggplot(data=x_p1_1,aes(x=year,group = 1))
g %>%
+geom_line(aes(y=total),linetype="solid",color="deepskyblue4") %>%
+geom_line(aes(y=fatal),linetype="solid",color="red") %>%
+geom_line(aes(y=non_fatal),linetype="solid",color="green") %>%
+theme_minimal() %>%
+theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) %>%
+labs(x = "Years",y="Total Health incidents", title = "Time series plot",subtitle = "breakdown of fatalities, injuries and total",colour= "Metric")

```


Evolve of how Health Injuries evolve over time (Year)



2. Across the United States, which types of events have the greatest economic consequences?

This requires a little more of transformation, we part from the "matchwith" column which 
we have been working with. 
Checking columns in the data, this two columns show economic damage on property and crops 
("PROPDMG","CROPDMG"), so we're going to adjust with the corresponding EXP columns to get 
a comparable value ("PROPDMGEXP","CROPDMGEXP").
Rules are:
```{r}
x_minimal <- x[,c("matchwith","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP")]
```

Existing values of exponents

```{r}
table(x_minimal$PROPDMGEXP)
table(x_minimal$CROPDMGEXP)
```

mapping and equivalencies
```{r}
v_exp   <-c("-","?","+","0","1","2","3","4","5","6","7","8","9","B","h","H","K","k","m","M")
v_value <-c(0,0,1,10,10,10,10,10,10,10,10,10,10,1000000000,100,100,1000,1000,1000000,1000000)
v_transform <- data.frame(v_exp,v_value)
x_minimal_transform <- merge(x_minimal,v_transform,by.x = c("PROPDMGEXP"),by.y = c("v_exp"),all.x = TRUE ) 
x_minimal_transform <- merge(x_minimal_transform,v_transform,by.x = c("CROPDMGEXP"),by.y = c("v_exp"),all.x = TRUE ) 

x_minimal_transform$CROPDMG_VALUE<-x_minimal_transform$CROPDMG*x_minimal_transform$v_value.y
x_minimal_transform$PROPDMG_VALUE<-x_minimal_transform$PROPDMG*x_minimal_transform$v_value.x
x_minimal_transform$total=coalesce(x_minimal_transform$PROPDMG_VALUE,0)+coalesce(x_minimal_transform$CROPDMG_VALUE,0)
```

Creating a dataframe grouped by event_type, and ordered by the damage value 

```{r}
x_p2 <-
x_minimal_transform %>%
  group_by(matchwith) %>%
  summarize( Total = sum(total))
x_p2 <- x_p2[order(-x_p2$Total),]
```

## Results

For the question 1
Here is the top 10 events wich are the more harmful by event type, total represent both fatalities and injuries.
```{r}
head(x_p1[,c('matchwith','total')],10)
```
For the question 2:
data.frame of top 10 evet_types most costly $ in the time period analysed, we skip null because of "summaries" those we  classified then as NA

```{r}
head(x_p2[complete.cases(x_p2),],10)
```



***
### Useful resources
> + [Storm data documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)  
> + [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)
> + [DataSource](https://www.ncdc.noaa.gov/stormevents/ftp.jsp)  
> + [Detailed information about the fields columns](https://www.ncei.noaa.gov/pub/data/swdi/stormevents/csvfiles/Storm-Data-Bulk-csv-Format.pdf)
> + [(1) Detailed information about the fields columns 2](https://www.ncei.noaa.gov/pub/data/swdi/stormevents/csvfiles/Storm-Data-Export-Format.pdf)
> + [Storm events data NOAA SOURCE](https://www.ncei.noaa.gov/pub/data/swdi/stormevents/csvfiles/)  
> + [R Markdown CheatSheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)  
> + [The stringdist package for approximate string matching](https://CRAN.R-project.org/package=stringdist)
> + [Getting he min distance of string comparison](https://stackoverflow.com/questions/50520702/extract-best-match-from-string-distance-matrix)
> + [Usana Foad - Coursera Help](https://www.coursera.org/learn/reproducible-research/discussions/weeks/4/threads/38y35MMiEeiERhLphT2-QA)
> + [Understanding PROPDMGEXP and CROPDMGEXP](https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html)
