---
title: "Final project Reroducible Research"
author: "Franco Israel Corona Hernandez"
date: "7/11/2021"
output: 
  html_document:
    keep_md: true
---

## Previus configs
```{r setup II, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(eval = TRUE, echo = TRUE)
#install.packages("dplyr")
#install.packages("stringdist")
library("dplyr")            ## load
library(stringdist)

```

  * Done


## Getting data
```{r gettingInicialFile}
final_file = "data.csv.bz2"
url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if(!file.exists(final_file)){
  download.file(url,final_file,quiet = FALSE,method = "curl")
}else{ 
  print(paste("File ",final_file,"already exists. Skipping download.",collapse = ""))
}

```
## Understanding data in file
 - Showing 10 instead of 1000 original lines loaded to keep more clean the output / report

```{r briefRaw}
file_raw <- readLines(con<-file(final_file),n=1000)
close(con)
head(file_raw,10)
```

This brief of file shows us that there is a comma separated file with headers, double quote as columns strings enclosure. 
Most of the data is numeric, at first sight 5 colulmns are posible string factors , it seems that null values are empty string, dates have date format 4/18/1950 and have a part of time in 0, time is in a INT format in 24 hrs format (1600)

## Loading and Processing the data

```{r loadDataR, cache=TRUE}
data_pp_1 <- read.table(file = final_file,sep = ",",header = TRUE,quote = "\"",na.strings = c("","NA"))

```

Loading seems correct now, a litte metrics in general about the dataset

```{r summary1}
summary(data_pp_1)
```

 + First values
 
```{r headFirstRows}

head(data_pp_1,10)

```

  + Last Values

```{r tailLastRows}
tail(data_pp_1,10)
```



Observations:  
  - Total Number of rows is **`r nrow(data_pp_1) `**, **`r ncol(data_pp_1) `** variables  
  - *F* column is almost empty on all the data.set  
  - *COUNTYENDN* is always NA  
  - *COUNTY_END* is always 0  
  - *Length*, *Width* are 0 at least at the 3rd quantile  
  - *BGN_TIME* has mixed values, integers and time strings  

* Using a different data.frame to keep the original for further checks

```{r dfClone}
data_pp_2<-data.frame()
data_pp_2<-data_pp_1
```

* Changing data types of dates: BGN_DATE,END_DATE

```{r transforms}

data_pp_2$BGN_DATE <- as.Date(data_pp_2$BGN_DATE,format="%m/%d/%Y %H:%M:%S")
data_pp_2$END_DATE <- as.Date(data_pp_2$END_DATE,format="%m/%d/%Y %H:%M:%S")
summary(data_pp_2[,c("BGN_DATE","END_DATE")])

```

## Questions

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?


```{r eval=FALSE}
#length(table(data_pp_2$STATE__))70?
#dimnames(table(data_pp_2$EVTYPE))

```

First we are going to analyse the EVTYPE column

```{r eval=TRUE, include=TRUE}
#evt<-(table(data_pp_2$EVTYPE))
#evt[1]
evt <-
  data_pp_2 %>%
  group_by(EVTYPE) %>%
  summarize(total = n())
#evt
head(evt[order(-evt$total),],10)

```

Because there are `r length(evt)` different events, I decided that we could use some help (1). Getting a cleaner catalog of events, therefore we are going to match this by string distance. 

First some cleaning and data wrangling

```{r}
event_type_catalog <- read.csv2(file = "EVTYPE.csv",header = TRUE)
evt$leandata <- evt$EVTYPE
evt$leandata <- as.character(evt$leandata)
evt$leandata <- trimws(evt$leandata)
evt$leandata <- tolower(evt$leandata)
event_type_catalog$EVTYPE <- as.character(event_type_catalog$EVTYPE)
event_type_catalog$EVTYPE <- trimws(event_type_catalog$EVTYPE)
event_type_catalog$EVTYPE <- tolower(event_type_catalog$EVTYPE)
```

Quick look of the data.frames

```{r}
head(evt,10)
head(event_type_catalog,10)
```
Hoe many rows by event type the original data.frame has, quick view

```{r eval=TRUE, include=TRUE}
evt2 <- 
  evt %>%
  group_by(leandata) %>%
  summarize(total = sum(total))

head(evt2[order(-evt2$total),],10)
```

Starting the process of getting the distances by jaro-winker method (closer to 0 is better)

```{r}

# Reclassification of EVTYPE to get a more accurate global classification

# Actual eventtype (dirty data)
uniquemodels <- unique(as.character(evt2$leandata))
# Catalog of NOAA to match
uniquemodels2 <- unique(as.character(event_type_catalog$EVTYPE)) 

# Calculation of the string distances by Jaro–Winkler
distancemodels <- stringdistmatrix(uniquemodels2,uniquemodels,method = "jw")

# setting colnames and rownames to identify which row and column correspond to each one distance
rownames(distancemodels) <- uniquemodels2
colnames(distancemodels) <- uniquemodels
#hc <- hclust(as.dist(distancemodels,upper = TRUE))
#plot(hc)
#rect.hclust(hc,k=20)
#list of minimun values (minimum values per column)
mins<-apply(distancemodels, 2, min)
#return list of the row number of the minimum value (most similar row vs columns)
minlist<-apply(distancemodels, 2, which.min)
#return list of matching values, minlist are the values of the NOAA catalog with is more likely to match the dirty data
matchwith<-uniquemodels2[minlist]

answer<-data.frame(uniquemodels, matchwith, mins)
maxDist<-0.30
answer_filter <- answer[(answer$mins<=maxDist)&(order(-answer$mins)),]
head(answer_filter[order(-answer_filter$mins),],10)

```
Getting in the **evt2** data.frame the values adjusted to the master catalog of event_types
```{r}
evt2_match <- merge(evt2,answer_filter,by.x=c("leandata"),by.y=c("uniquemodels"),all.x = TRUE)
head(evt2_match[order(-evt2_match$total),],10)
```

Lets do a quick analysis of the percentiles in where the data is null, separate the data in null and not null

```{r}
evt2_match$factor_matchwith<-0
evt2_match[complete.cases(evt2_match),'factor_matchwith'] <- 1

```


```{r}
evt3 <- 
  evt2_match %>%
  group_by(factor_matchwith) %>%
  summarize(total = sum(total), total_cn = n())

evt3[order(-evt3$total),]
```
At a **maxDist** of .30 we found a good proportion of data events  re-classified, `r (evt3[evt3$factor_matchwith==0,'total']/evt3[evt3$factor_matchwith==1,'total'])*100`% are misspelled events, summary or something that occur one time and not relevant


2. Across the United States, which types of events have the greatest economic consequences?


## Results

***
### Useful resources
> [DataSource](https://www.ncdc.noaa.gov/stormevents/ftp.jsp)  
> + [Detailed information about the fields columns](https://www.ncei.noaa.gov/pub/data/swdi/stormevents/csvfiles/Storm-Data-Bulk-csv-Format.pdf)
> + [(1) Detailed information about the fields columns 2](https://www.ncei.noaa.gov/pub/data/swdi/stormevents/csvfiles/Storm-Data-Export-Format.pdf)
> + [Documentation on the file naming convention](http://www1.ncei.noaa.gov/pub/data/swdi/stormevents/csvfiles/README)  
> + [Storm events data NOAA](https://www.ncei.noaa.gov/pub/data/swdi/stormevents/csvfiles/)  
> + [R Markdown CheatSheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)  
> + [The stringdist package for approximate string matching](https://CRAN.R-project.org/package=stringdist)
> + [Getting he min distance of string comparison](https://stackoverflow.com/questions/50520702/extract-best-match-from-string-distance-matrix)
> + [Usana Foad - Coursera Help](https://www.coursera.org/learn/reproducible-research/discussions/weeks/4/threads/38y35MMiEeiERhLphT2-QA)
> + [Understanding PROPDMGEXP and CROPDMGEXP](https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html)
